### 2022-09-28 main/chapter_context.tex
write some on origins

\subsection{Opportunities for future Standard Models}
Problems with the Standard Model mean it needs replacing with supersymmetry
--- is a false statement.
Such an argument could motivate this thesis, but the Standard Model is actually
good in its domain.
This discussion therefore looks towards where an altered Standard Model could
go to improve its current abilities or to expand its generality.

When `nobody knows' the answer to a riddle, the status lies somewhere between
two extremes:
either many answers appear satisfactory and none is the clear best,
or no satisfactory answer is yet known.
We distinguish these two cases with the respective labels
\emph{uncertain} and \emph{unknown}:
the model of neutrino masses is mostly \emph{uncertain}, but
the reconciliation quantum field theory with general relativity is
mostly \emph{unknown}.

One problem with the Standard Model is that algorithms to compute its
\emph{general} consequences are unknown
where perturbative and discrete lattice methods are impractical.
But this is not an opportunity because it is inherent to the framework
of quantum field theory, which all competitive alternatives also use.
Before introducing supersymmetry, this section reviews some promising
directions of change for future models.

\paragraph{Neutrino masses.}
Neutrinos could be Dirac fermions with Yukawa masses and boring right-chiral
fields.
Or they could also have Majorana mass terms, for which the seesaw mechanism
can simultaneously generate the observed light neutrinos and heavy `sterile'
neutrinos.
Or their masses could come from other operators without right-handed
fields~\cite{
thomson2013modern,
chala2021neutrino,
wells2020discovery
}.
The answer is uncertain, and new results from theory or experiment that
pinpoint a correct model could improve the Standard Model in the neutrino
domain.

@@

The universe contains dark matter which interacts with gravity but not much
with other forces; it is observed through galactic rotation curves,
gravitational lensing, and cosmic microwave background data,
but its composition is uncertain~\cite{
begeman1991rotation,
garrett2010dark,
planck2020results
}.
Dark matter candidates include new particles such as
sterile neutrinos~\cite{boyarsky2019sterile},
WIMPs (Weakly Interacting Massive Particles)~\cite{jungman1996wimp},
or
axions~\cite{kim2008axions},
but also macroscopic objects such as
primordial black holes~\cite{bernard2019primordial}.
Dark matter may well comprise various of different components, and it would
be a success for future particle physics theories to pinpoint necessary dark
matter particles.

Other mysteries of cosmology include:
the positive cosmological constant,
the inflation of the early universe,
and baryogenesis~\cite{
wells2020discovery,
riess1998observational
}.
Each lies somewhere between uncertain and unknown, but may not be entirely
out of reach;
the Higgs field could plausibly have driven
inflation~\cite{bezrukov2008higgs},
and the Standard Model does slightly violate $C$ and $CP$, both of which
are required for baryogenesis~\cite{sakharov1991re}.

\paragraph{Origins.}
\TODO{parameters and structure}

\paragraph{Aesthetics.}

\TODO{naturalness and fine-tuning~\cite{wells2020discovery}}
\TODO{strong CP problem}
\TODO{hierarchy problem}

@@

% using https://www.tablesgenerator.com/latex_tables
\begin{table}[tp]
\centering
\begin{tabular}{rrc}
\multirow{5}{*}{bosons}   & \multirow{2}{*}{squarks}      & \makebox[5em]{\hfill $\tilde u$ \hfill $\tilde c$ \hfill $\tilde t$ \hfill}                  \\
                          &                               & \makebox[5em]{\hfill $\tilde d$ \hfill $\tilde s$ \hfill $\tilde b$ \hfill}                  \\[1ex]
                          & \multirow{2}{*}{sleptions}    & \makebox[5em]{\hfill $\tilde e$ \hfill $\tilde \mu$ \hfill $\tilde \tau$ \hfill}             \\
                          &                               & \makebox[5em]{\hfill $\tilde \nu_e$ \hfill $\tilde \nu_\mu$ \hfill $\tilde \nu_\tau$ \hfill} \\[1ex]
                          & Higgs                         & \makebox[7em]{\hfill ($h$) \hfill $H^0$ \hfill $A^0$ \hfill $H^\pm\!\!$ \hfill}             \\[3ex]
\multirow{3}{*}{fermions} & strong                        & \makebox[5em]{\hfill $\tilde g$ \hfill}                                        \\[1ex]
                          & \multirow{2}{*}{electroweak}  & \makebox[5em]{\hfill $\tilde C^\pm_1\!\!$ \hfill $\tilde C^\pm_2\!\!$ \hfill} \\
                          &                               & \makebox[7em]{\hfill $\tilde N_1\!$ \hfill $\tilde N_2\!$ \hfill$\tilde N_3\!$ \hfill$\tilde N_4\!$ \hfill} \\
\end{tabular}
\caption[Particles of the Minimal Supersymmetric Standard Model]{%
Particles of the Minimal Supersymmetric Standard Model
\TODO{relate to Standard Model}
\TODO{adds goldstino and gravitino}
\TODO{alternative $\chargino_i = \tilde C^\pm_i$, $\neutralino_i = N_i$}
\TODO{In this table, all bosons are spin-$0$ scalars and all fermions are spin-$1/2$.}
\TODO{The bracketed $h$ corresponds to the Standard Model Higgs boson.}
}
\label{tab:context_particles_mssm}
\end{table}

>>>>

\subsection{Opportunities for future Standard Models}
Problems with the Standard Model mean it needs replacing with supersymmetry
--- is a false statement.
Such an argument could motivate this thesis, but the Standard Model is actually
good in its domain.
This discussion therefore looks towards where an altered Standard Model could
improve its current abilities or to expand its generality to describe more of
the universe.

When `nobody knows' the answer to a riddle, the status lies somewhere between
two extremes:
either many answers are satisfactory with no clear best,
or no satisfactory answers are yet known.
We distinguish these two cases with the labels
\emph{uncertain} and \emph{unknown} respectively:
the model of neutrino masses is mostly \emph{uncertain}, but
the reconciliation quantum field theory with general relativity is
mostly \emph{unknown}.

One real problem with the Standard Model is that algorithms to compute its
general consequences are unknown, since practical perturbative and discrete
lattice methods are limited.
Such imprecise modelling weakens our inferences from experimental results,
anomalous or not,
but does not present much opportunity because it is common to the quantum field
theory that is used by all known viable models.
This section briefly reviews some other directions of positive change for
future models.

\paragraph{Neutrino masses.}
Neutrinos could be Dirac fermions with Yukawa masses and boring right-chiral
fields.
They could also have Majorana mass terms, for which the seesaw mechanism can
simultaneously generate the observed light neutrinos with very heavy `sterile'
neutrinos.
Or their masses could come from other operators without needing the
right-chiral fields~\cite{
thomson2013modern,
chala2021neutrino,
wells2020discovery
}.
This answer is uncertain, and new results from theory or experiment that
pinpoint a correct model could improve the Standard Model in the neutrino
domain.

@@

The universe contains dark matter which interacts with gravity but not much
with other forces; it is observed through galactic rotation curves,
gravitational lensing, and cosmic microwave background data,
but its composition is uncertain~\cite{
begeman1991rotation,
garrett2010dark,
planck2020results
}.
Dark matter candidates include new particles such as
sterile neutrinos~\cite{boyarsky2019sterile},
Weakly Interacting Massive Particles (WIMPs)~\cite{jungman1996wimp},
or
axion-like particles~\cite{kim2008axions},
but also macroscopic objects such as
primordial black holes~\cite{bernard2019primordial}.
Dark matter may well comprise various of different components, and it would
be a success for future particle physics theories to pinpoint necessary dark
matter particles.

Other mysteries of cosmology include:
the positive cosmological constant,
the inflation of the early universe,
and baryogenesis~\cite{
wells2020discovery,
riess1998observational
}.
Answers to these mysteries all lie somewhere between uncertain and unknown,
but may not be entirely out of reach;
for example, the Higgs field could plausibly have driven
inflation~\cite{bezrukov2008higgs},
and the Standard Model does slightly violate $C$ and $CP$, both of which
are required for baryogenesis~\cite{sakharov1991re}.

\paragraph{Origins.}
Why does the Standard Model have its structure?
Why is it configured with the observed values of our parameters?
Why do we even live in $3+1$ dimensional spacetime with quantum mechanics?
Answers to these questions are largely unknown, but would be
attractive features for a fundamental theory.

Certain Standard Model structures are quite general and parametrized such
that parts can be tuned to describe nature.
For example, the strong interaction includes a parameter $\theta_\mathrm{CP}$ whose
non-zero values would induce $\mathrm{CP}$ violation.
Observations find it very close to zero, suggesting that nature does not use
that freedom, so one promising feature of axion-like models is their
implication that $\theta_\mathrm{CP} = 0$~\cite{
kim2008axions,
thomson2013modern,
martin2017particle
}.

Justified constraints on the model could have led to better predictions of
past data and might extrapolate to new phenomena through their deeper
understanding of nature.
Attempts to relate empirical parameters~\cite{
sato1979ratio,
beg1982gauge,
koide1983fermion,
denterria2012gaussian
}
or to derive mathematical structures~\cite{
goyal2010origin,
skilling2021abc,
axioms1010038
}
are curious, but will remain speculative until they provide
real and novel benefits.

\paragraph{Aesthetics.}
\TODO{naturalness and fine-tuning~\cite{wells2020discovery}}
\TODO{strong CP problem}
\TODO{hierarchy problem}

Naturalness is opinion and only nature decides what is really natural.

@@

% using https://www.tablesgenerator.com/latex_tables
\begin{table}[tp]
\centering
\begin{tabular}{rrc}
\multirow{5}{*}{bosons}   & \multirow{2}{*}{squarks}      & \makebox[5em]{\hfill $\tilde u$ \hfill $\tilde c$ \hfill $\tilde t$ \hfill}                  \\
                          &                               & \makebox[5em]{\hfill $\tilde d$ \hfill $\tilde s$ \hfill $\tilde b$ \hfill}                  \\[1ex]
                          & \multirow{2}{*}{sleptions}    & \makebox[5em]{\hfill $\tilde e$ \hfill $\tilde \mu$ \hfill $\tilde \tau$ \hfill}             \\
                          &                               & \makebox[5em]{\hfill $\tilde \nu_e$ \hfill $\tilde \nu_\mu$ \hfill $\tilde \nu_\tau$ \hfill} \\[1ex]
                          & Higgs                         & \makebox[7em]{\hfill ($h$) \hfill $H^0$ \hfill $A^0$ \hfill $H^\pm\!\!$ \hfill}             \\[3ex]
\multirow{3}{*}{fermions} & strong                        & \makebox[5em]{\hfill $\tilde g$ \hfill}                                        \\[1ex]
                          & \multirow{2}{*}{electroweak}  & \makebox[5em]{\hfill $\tilde C^\pm_1\!\!$ \hfill $\tilde C^\pm_2\!\!$ \hfill} \\
                          &                               & \makebox[7em]{\hfill $\tilde N_1\!$ \hfill $\tilde N_2\!$ \hfill$\tilde N_3\!$ \hfill$\tilde N_4\!$ \hfill} \\
\end{tabular}
\caption[Particles of the Minimal Supersymmetric Standard Model]{%
Particles of the Minimal Supersymmetric Standard Model
\TODO{check and cite}
\TODO{relate to Standard Model}
\TODO{adds goldstino and gravitino}
\TODO{alternative $\chargino_i = \tilde C^\pm_i$, $\neutralino_i = N_i$}
\TODO{In this table, all bosons are spin-$0$ scalars and all fermions are spin-$1/2$.}
\TODO{The bracketed $h$ corresponds to the Standard Model Higgs boson.}
}
\label{tab:context_particles_mssm}
\end{table}
